---
title: "R code for Data Science for Beginners"
subtitle: " Day 5: Individual Exercise"
author: "Graham Jones"  
date: "2025-09-20"  
format: pdf
---

# Clean up your workspace 
```{r}
rm(list=ls(all=TRUE)) # delete all objects in the environment 
cat("\014") # clear the console screen
library(tidyverse)
```

# Load the data 
```{r}
world_data = read.csv("~/Downloads/world.csv")
```

Remember that Quarto uses a relative path so always save your data in the same folder (or under the same folder) with your Quarto code. 

## Democracy and female representation 

Do democratic countries (`democ_regime == "Yes"`) have better female representation than non-democratic countries? Please answer this question by showing some graphs to assess the relationship between these two variables. 


### Initial data clean up

Analyze y
```{r}
summary(world_data $ women09) # look at summary statistics for female representation
```

Analyze x
```{r}
summary(world_data $ democ_regime) # look at summary of regime type (Yes/No)
```

Note: There are some missing values. Labels are not intuitive. So, we will deal with these two first.


### Create a smaller data set that omits NA observations.
```{r}
women_data <- world_data[ is.na(world_data $ women09) == FALSE & 
                           is.na(world_data $ democ_regime) == FALSE, ] # keep only rows without missing values
```

Let's make sure that we did this correctly. 

```{r}
summary(women_data $ women09) # check again, no missing values
summary(women_data $ democ_regime)  # check regime variable again
```

We can see that NA cases have been correctly removed. 


### Re-labeling the values
```{r}
women_data $ regime_label <- factor(women_data $ democ_regime, 
                              levels = c("Yes", "No"),
                              labels = c("Democracy", "Autocracy"))

summary(women_data $ regime_label) # check the new labels
rm(world_data) # The original dataset is no longer needed.
```

## Q1: Describe Y (numerically and graphically) 

### Numerical summary


```{r}
summary(women_data $ women09) # numerical summary of female representation
```


### Graphical summary


```{r}
ggplot(women_data, aes(x=women09)) +
  geom_histogram(bins = 30) # histogram of female representation
```


## Q2: Describe X (numerically and graphically) 


### Numerical summary (frequency table)


```{r}
summary(women_data$regime_label) # frequency count of regime type
```


### Graphical summary (bar chart)


```{r}
ggplot(women_data, aes(x = regime_label)) +
  geom_bar()  # bar chart of countries by regime type
```


## Q3: Describe X-Y graphically

### Histograms


```{r}
ggplot(women_data, aes(women09))+
  geom_histogram(bins=30) +
  theme(axis.text.x = element_text(size = 14)) +
  xlab("Fem Representation") +
  ylab("Number of Countries") +
  theme_bw() +
  facet_grid(.~regime_label) # histograms split by regime type
```

### Box-plots


```{r}
ggplot(women_data, aes(x = regime_label, y = women09))+
  geom_boxplot() +
  xlab("Political System") +
  ylab("Fem Representation") # boxplot comparing democracies and autocracies
```

