---
title: "HW12"
author: "Graham Jones"
date: "November 10, 2025"
format:
  pdf:
    number-sections: false
    toc: false
    pdf-engine: xelatex
    fig-cap-location: bottom
    fig-align: center
    colorlinks: true
    linkcolor: RoyalBlue
    urlcolor: RoyalBlue
fontsize: 11pt
geometry: margin=1in
highlight-style: tango
execute:
  echo: true
  warning: false
  message: false
  error: true
code-fold: true
code-summary: "Show code"
code-overflow: wrap
editor: visual
---

# Load setup and themes
# I began by loading the required packages and setting consistent figure styles.
```{r}
suppressPackageStartupMessages({
  library(effects)
  library(ggplot2)
})
knitr::opts_chunk$set(fig.width = 6.5, fig.height = 4.2, dpi = 300, fig.align = "center")
theme_set(
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
)
```
# Load dataset
# Here I import the dataset from the provided Dropbox link.
```{r}
world <- read.csv("https://www.dropbox.com/scl/fi/1dzteukcseiy51pjbgf0b/world.csv?rlkey=pm9u4jcoszwyy4awlxulc0nwa&e=1&st=msttbjgs&dl=1")
```
# Clean pr_sys variable
# I ensure that the electoral system variable is labeled properly as “Non-PR” and “PR.”
```{r}
if (!("pr_sys" %in% names(world))) stop("Column 'pr_sys' not found in dataset.")
if (!is.factor(world$pr_sys)) {
  vals <- sort(unique(na.omit(world$pr_sys)))
  if (all(vals %in% c(0, 1))) {
    world$pr_sys <- factor(world$pr_sys, levels = c(0, 1), labels = c("Non-PR", "PR"))
  } else {
    world$pr_sys <- factor(world$pr_sys)
  }
}
print(paste0("pr_sys levels: ", paste(levels(world$pr_sys), collapse = ", ")))
```
# Q1) Constant-only model
# I start with a constant-only model to find the mean female representation.
```{r}
model1 <- lm(women09 ~ 1, data = world)
summary(model1)
```
# Q2) Model with per capita GDP
# I add GDP per capita to test if wealthier countries have higher representation.
```{r}
model2 <- lm(women09 ~ gdp_10_thou, data = world)
summary(model2)
```
# Q3) Graph: Effect of per capita GDP
# I visualize how GDP affects female representation with confidence bands.
```{r}
df_eff2 <- as.data.frame(effects::Effect("gdp_10_thou", model2))
ggplot(df_eff2, aes(x = gdp_10_thou, y = fit)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18) +
  geom_line(linewidth = 0.9) +
  labs(
    title = "Effect of Per Capita GDP on Female Representation",
    x = "GDP (in $10,000s)",
    y = "Predicted Female Representation (%)"
  )
```
# Q4) Model with electoral system dummy
# I now include electoral system type to test its independent effect.
```{r}
model3 <- lm(women09 ~ pr_sys, data = world)
summary(model3)
```
# Q5) Graph: Effect of electoral system
# I plot predicted values by system type.
```{r}
df_eff3 <- as.data.frame(effects::Effect("pr_sys", model3))
ggplot(df_eff3, aes(x = pr_sys, y = fit)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  labs(
    title = "Effect of Electoral System on Female Representation",
    x = "Electoral System",
    y = "Predicted Female Representation (%)"
  )
```
# Q6) Model with BOTH GDP and electoral system
# I combine both predictors to see how they behave together.
```{r}
model4 <- lm(women09 ~ gdp_10_thou + pr_sys, data = world)
summary(model4)
```
# Q7) Compare model fit (R² and AIC)
# I compare model performance to see which fits best.
```{r}
print(c(
  R2_m1 = summary(model1)$r.squared,
  R2_m2 = summary(model2)$r.squared,
  R2_m3 = summary(model3)$r.squared,
  R2_m4 = summary(model4)$r.squared
))
AIC(model1, model2, model3, model4)
```
# Q8) Graph: GDP effect for PR countries only
# I show the relationship between GDP and representation within PR systems.
```{r}
target_level <- if ("PR" %in% levels(world$pr_sys)) "PR" else levels(world$pr_sys)[1]
if (!("PR" %in% levels(world$pr_sys))) {
  print(paste0("Note: 'PR' not found. Using level '", target_level, "' instead."))
}
gseq <- seq(min(world$gdp_10_thou, na.rm = TRUE),
            max(world$gdp_10_thou, na.rm = TRUE),
            length.out = 100)
nd_pr <- data.frame(
  gdp_10_thou = gseq,
  pr_sys = factor(target_level, levels = levels(world$pr_sys))
)
pred_pr <- predict(model4, newdata = nd_pr, se.fit = TRUE)
nd_pr$fit <- pred_pr$fit
nd_pr$lwr <- pred_pr$fit - 1.96 * pred_pr$se.fit
nd_pr$upr <- pred_pr$fit + 1.96 * pred_pr$se.fit
ggplot(nd_pr, aes(x = gdp_10_thou, y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.18) +
  geom_line(linewidth = 0.9) +
  labs(
    title = paste0("Effect of GDP on Female Representation (", target_level, " Only)"),
    x = "GDP (in $10,000s)",
    y = "Predicted Female Representation (%)"
  )
```
# Q9) Interaction (GDP × PR)
# I test for an interaction to see if GDP’s impact differs between PR and Non-PR systems.
```{r}
model_interact <- lm(women09 ~ gdp_10_thou * pr_sys, data = world)
gseq <- seq(min(world$gdp_10_thou, na.rm = TRUE),
            max(world$gdp_10_thou, na.rm = TRUE),
            length.out = 100)
nd_all <- expand.grid(
  gdp_10_thou = gseq,
  pr_sys = levels(world$pr_sys)
)
pred_all <- predict(model_interact, newdata = nd_all, se.fit = TRUE)
nd_all$fit <- pred_all$fit
nd_all$lwr <- nd_all$fit - 1.96 * pred_all$se.fit
nd_all$upr <- nd_all$fit + 1.96 * pred_all$se.fit
ggplot(nd_all, aes(x = gdp_10_thou, y = fit, color = pr_sys, fill = pr_sys)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.12, linewidth = 0) +
  geom_line(linewidth = 0.9) +
  labs(
    title = "Effect of GDP by Electoral System",
    x = "GDP (in $10,000s)",
    y = "Predicted Female Representation (%)",
    color = "System", fill = "System"
  )
```
# Q10) Region model
# I now test for regional variation in women’s representation.
```{r}
model_region <- lm(women09 ~ region, data = world)
summary(model_region)
```
# Q11) Graph: Female representation by region
# I plot predicted averages by region.
```{r}
df_eff_region <- as.data.frame(effects::Effect("region", model_region))
ggplot(df_eff_region, aes(x = region, y = fit)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  labs(
    title = "Female Representation by Region",
    x = "Region",
    y = "Predicted Female Representation (%)"
  )
```
# Q12) Model: GDP controlling for region
# I control for GDP to isolate its impact within regional context.
```{r}
model_gdp_region <- lm(women09 ~ gdp_10_thou + region, data = world)
summary(model_gdp_region)
```
# Clean frac_eth3 variable
# Before modeling ethnic fractionalization, I clean missing and blank entries.
```{r}
if (!"frac_eth3" %in% names(world)) stop("Column 'frac_eth3' not found in dataset.")
KEEP_MISSING  <- FALSE
MISSING_LABEL <- "Missing"
x <- as.character(world$frac_eth3)
is_blank <- !is.na(x) & nchar(trimws(x)) == 0
x[is_blank] <- NA_character_
if (KEEP_MISSING) {
  x[is.na(x)] <- MISSING_LABEL
  world$frac_eth3 <- factor(x)
  print("Kept missing/blank values as labeled.")
} else {
  keep <- !is.na(x)
  dropped <- sum(!keep)
  world <- world[keep, , drop = FALSE]
  world$frac_eth3 <- factor(x[keep])
  print(paste("Dropped", dropped, "rows with missing/blank frac_eth3."))
}
print(paste("Final frac_eth3 levels:", paste(levels(world$frac_eth3), collapse = ", ")))
```
# Q13) Ethnic fractionalization model
# I estimate representation across ethnic fractionalization categories.
```{r}
model_ethnic <- lm(women09 ~ frac_eth3, data = world)
summary(model_ethnic)
```
# Q14) Graph: Effect of ethnic fractionalization
# I visualize predicted representation by fractionalization level.
```{r}
eff_eth <- effects::Effect("frac_eth3", model_ethnic)
df_eff_eth <- as.data.frame(eff_eth)
ggplot(df_eff_eth, aes(x = frac_eth3, y = fit)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  geom_text(aes(label = sprintf("%.1f", fit)), vjust = -0.5, size = 3.3) +
  labs(
    title = "Effect of Ethnic Fractionalization on Female Representation",
    x = "Ethnic Fractionalization Level",
    y = "Predicted Female Representation (%)"
  ) +
  ylim(0, max(df_eff_eth$upper, na.rm = TRUE) * 1.1)
```












