---
title: "POLI502 Final Exam"
author: "Graham Jones"
date: "2025-12-05"
output: pdf_document
---

```{r setup, include=FALSE}
#Set global chunk options and load packages
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
library(tidyverse)
library(stargazer)
library(pROC)
library(ggplot2)
```

# Question 1: Exploring Fare by Gender

```{r q1}
# Load Titanic dataset
td <- read.csv("titanic2.csv")

# Ensure female is treated as a factor
td$female <- factor(td$female)

# Explore relationship between female and fare
# Summarize fares by gender: mean, median, standard deviation, and count
fare_by_gender <- td %>%
  group_by(female) %>%
  summarise(
    mean_fare = mean(fare, na.rm = TRUE),
    median_fare = median(fare, na.rm = TRUE),
    sd_fare = sd(fare, na.rm = TRUE),
    n = n()
  )
# Show the summary table
print(fare_by_gender)

# Draw a boxplot to compare fare distributions by gender
ggplot(td, aes(x = female, y = fare, fill = female)) +
  geom_boxplot() +
  labs(title = "Distribution of Ticket Fare by Gender",
       x = "Gender", y = "Fare (GBP)") +
  theme_minimal()

# Female passengers appear to have higher median and mean fares
# compared to male passengers, suggesting they tended to purchase more expensive tickets.
```

# Question 2: Bivariate Statistical Test

```{r q2}
# Perform t-test to compare fares between genders
t_test_result <- t.test(fare ~ female, data = td)
# Display the t-test result
print(t_test_result)

# The t-test shows that female passengers had significantly higher ticket fares than male 
# passengers (p < 0.05). The mean fare for female passengers was approximately £44.5 
# compared to £25.5 for male passengers. Therefore, we can conclude that female passengers 
# tended to have more expensive tickets than male passengers.
```

# Question 3a: Logit Models

```{r q3a}
# Estimate logit model with original fare as a predictor
model1 <- glm(survived ~ fare + female + child, 
              data = td, family = binomial(link = "logit"))

# Estimate a second logit model using log(fare + 1) instead of fare
model2 <- glm(survived ~ I(log(fare + 1)) + female + child, 
              data = td, family = binomial(link = "logit"))

# Show both models in a stargazer table
stargazer(model1, model2, type = "text",
          title = "Logit Models of Passenger Survival",
          column.labels = c("Original Fare", "Log Fare"),
          covariate.labels = c("Fare", "Log(Fare+1)", "Female", "Child"),
          dep.var.labels = "Survived")
```

# Question 3b: Model Comparison

```{r q3b}
# Compare AIC values of the two models (lower AIC = better fit)
aic_comparison <- data.frame(
  Model = c("Model 1 (Original Fare)", "Model 2 (Log Fare)"),
  AIC = c(AIC(model1), AIC(model2))
)
# Show the AIC comparison
print(aic_comparison)

# Model 2 with log(fare) performs better because it has a lower AIC value.
```

# Question 3c: Effect Graphs

```{r q3c}
# Find the most common value of child to hold constant (mode)
child_mode <- names(sort(table(td$child), decreasing = TRUE))[1]

# Create prediction data for Model 1 (vary fare from 0 to 500)
pred_data1 <- data.frame(
  fare   = seq(0, 500, length.out = 100),
  # Hold gender constant at "Female"
  female = factor("Female", levels = levels(td$female)),
  # Hold child at its most common category (character, matching the model)
  child  = child_mode
)
# Get predicted survival probabilities from Model 1
pred_data1$predicted_prob <- predict(model1, newdata = pred_data1, type = "response")

# Plot the effect of fare on survival (original fare, Model 1)
ggplot(pred_data1, aes(x = fare, y = predicted_prob)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Effect of Fare on Survival (Original Fare)",
       x = "Fare (GBP)", y = "Predicted Probability of Survival") +
  theme_minimal()

# Create prediction data for Model 2 (same fare range and constants)
pred_data2 <- data.frame(
  fare   = seq(0, 500, length.out = 100),
  female = factor("Female", levels = levels(td$female)),
  child  = child_mode
)
# Get predicted survival probabilities from Model 2
pred_data2$predicted_prob <- predict(model2,
                                     newdata = pred_data2,
                                     type = "response")

# Plot the effect of fare on survival (log fare model, Model 2)
ggplot(pred_data2, aes(x = fare, y = predicted_prob)) +
  geom_line(color = "red", size = 1) +
  labs(title = "Effect of Fare on Survival (Log Fare)",
       x = "Fare (GBP)", y = "Predicted Probability of Survival") +
  theme_minimal()
```

# Question 3d: Graph Comparison

```{r q3d}
# The graph with the original fare variable suggests that the effect of fare on
# survival is non-linear. At lower fare values, the predicted probability of
# survival increases rapidly as fare rises, but this increase slows down at higher
# fare levels, indicating diminishing returns. This pattern indicates that the
# marginal benefit of paying more for a ticket becomes smaller at the high end of
# the fare distribution. In contrast, the graph based on the logged fare variable
# shows a more linear relationship between fare and survival, with a steadier
# increase in predicted survival probabilities across the fare range. Taken
# together with the model fit statistics, these patterns make the logged-fare
# specification more plausible as a description of how ticket price relates to
# survival chances.
```

# Question 3e: Training and Test Sets

```{r q3e}
# Remove rows with missing fare so log(fare) is well-defined
td_complete <- td %>% filter(!is.na(fare))

# Set seed so the random split is reproducible
set.seed(123)

# Create training and test sets (80/20 split)
train_id <- sample(1:nrow(td_complete), nrow(td_complete) * 0.8)
train_data <- td_complete[train_id, ]
test_data <- td_complete[-train_id, ]

# Estimate model without logged fare (only female and child)
model_no_fare <- glm(survived ~ female + child, 
                     data = train_data, family = binomial(link = "logit"))

# Estimate model with logged fare plus female and child on training data
model_with_fare <- glm(survived ~ I(log(fare + 1)) + female + child, 
                       data = train_data, family = binomial(link = "logit"))
```

# Question 3f: Stargazer Table for Training Models

```{r q3f}
# Show training-set models with and without log(fare) in a stargazer table (Q3f)
stargazer(model_no_fare, model_with_fare, type = "text",
          title = "Training Set Models",
          column.labels = c("Without Fare", "With Log Fare"),
          covariate.labels = c("Female", "Child", "Log(Fare+1)"),
          dep.var.labels = "Survived")
```

# Question 3g: ROC Curves

```{r q3g}
# Predicted probabilities on the test set for both models
pred_no_fare <- predict(model_no_fare, newdata = test_data, type = "response")
pred_with_fare <- predict(model_with_fare, newdata = test_data, type = "response")

# Create ROC objects (true outcome vs predicted probability)
roc_no_fare <- roc(test_data$survived, pred_no_fare)
roc_with_fare <- roc(test_data$survived, pred_with_fare)

# Plot ROC curve for model without fare
plot(roc_no_fare, col = "blue", main = "ROC Curves Comparison")
# Add ROC curve for model with log(fare)
plot(roc_with_fare, col = "red", add = TRUE)
# Add legend labeling each curve
legend("bottomright", legend = c("Without Fare", "With Log Fare"),
       col = c("blue", "red"), lwd = 2)
```

# Question 3h: AUC Scores

```{r q3h}
# Calculate and report AUC scores
auc_no_fare <- auc(roc_no_fare)
auc_with_fare <- auc(roc_with_fare)

#Print AUC scores to compare predictive accuracy
print(paste("AUC without fare:", auc_no_fare))
print(paste("AUC with log fare:", auc_with_fare))
```

# Question 3i: Model Performance Comparison

```{r q3i}
# The model with log(fare) performs better based on both the ROC curves and 
# AUC scores. The ROC curve for the model with fare is positioned higher and more to the 
# left, indicating better classification performance. Additionally, the AUC score is higher 
# for the model with (log)fare, confirming its superior predictive ability.
```

# Question 4: Putnam Dataset - Civic Community Models

```{r q4}
# Load Putnam dataset
putnam <- read.csv("putnam.csv")

# Create North dummy variable: 1 = North, 0 = South
putnam$North <- ifelse(putnam$NorthSouth == "North", 1, 0)

# Model 1: Simple linear regression
putnam_model1 <- lm(InstPerform ~ CivicCommunity, data = putnam)

# Model 2: Additive model with North-South control
putnam_model2 <- lm(InstPerform ~ CivicCommunity + North, data = putnam)

# Model 3: Interactive model with different slopes by region
putnam_model3 <- lm(InstPerform ~ CivicCommunity * North, data = putnam)

# Show all three civic community models in one stargazer table
stargazer(putnam_model1, putnam_model2, putnam_model3, type = "text",
          title = "Regression Models: Institutional Performance and Civic Community",
          column.labels = c("Simple", "Additive", "Interactive"),
          dep.var.labels = "Institutional Performance")
```

# Question 5: Graphs for Civic Community Models

```{r q5}
# 5a: Prediction data for Model 1 across CivicCommunity range
pred_data_p1 <- data.frame(
  CivicCommunity = seq(min(putnam$CivicCommunity, na.rm = TRUE),
                            max(putnam$CivicCommunity, na.rm = TRUE),
                            length.out = 100)
)
pred_data_p1$predicted <- predict(putnam_model1, newdata = pred_data_p1)
# Plot marginal effect of Civic Community (Model 1)
ggplot(pred_data_p1, aes(x = CivicCommunity, y = predicted)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Marginal Effect of Civic Community Index (Model 1)",
       x = "Civic Community Index", y = "Predicted Institutional Performance") +
  theme_minimal()

# 5b: Graphs for Model 2 (separate for North and South)
pred_data_p2_north <- data.frame(
  CivicCommunity = seq(min(putnam$CivicCommunity, na.rm = TRUE),
                            max(putnam$CivicCommunity, na.rm = TRUE),
                            length.out = 100),
  North = 1
)
pred_data_p2_north$predicted <- predict(putnam_model2, newdata = pred_data_p2_north)
#Prediction data for Model 2, North = 0
pred_data_p2_south <- data.frame(
  CivicCommunity = seq(min(putnam$CivicCommunity, na.rm = TRUE),
                            max(putnam$CivicCommunity, na.rm = TRUE),
                            length.out = 100),
  North = 0
)
pred_data_p2_south$predicted <- predict(putnam_model2, newdata = pred_data_p2_south)
# Plot marginal effect for Northern regions (Model 2)
ggplot() +
  geom_line(data = pred_data_p2_north, aes(x = CivicCommunity, y = predicted), 
            color = "blue", size = 1) +
  labs(title = "Marginal Effect of Civic Community Index - Northern Regions (Model 2)",
       x = "Civic Community Index", y = "Predicted Institutional Performance") +
  theme_minimal()
# Plot marginal effect for Southern regions (Model 2)
ggplot() +
  geom_line(data = pred_data_p2_south, aes(x = CivicCommunity, y = predicted), 
            color = "red", size = 1) +
  labs(title = "Marginal Effect of Civic Community Index - Southern Regions (Model 2)",
       x = "Civic Community Index", y = "Predicted Institutional Performance") +
  theme_minimal()

# 5c: Prediction data for Model 3 over CivicCommunity and North
pred_data_p3 <- expand.grid(
  CivicCommunity = seq(min(putnam$CivicCommunity, na.rm = TRUE),
                            max(putnam$CivicCommunity, na.rm = TRUE),
                            length.out = 100),
  North = c(0, 1)
)
pred_data_p3$predicted <- predict(putnam_model3, newdata = pred_data_p3)
# Label each row as "North" or "South" for plotting
pred_data_p3$Region <- ifelse(pred_data_p3$North == 1, "North", "South")
# Plot marginal effects by region (two panels, Model 3)
ggplot(pred_data_p3, aes(x = CivicCommunity, y = predicted, color = Region)) +
  geom_line(size = 1) +
  facet_wrap(~Region) +
  labs(title = "Marginal Effect of Civic Community Index by Region (Model 3)",
       x = "Civic Community Index", y = "Predicted Institutional Performance") +
  theme_minimal()
```

# Question 6: Spuriousness Analysis - Civic Community

```{r q6}
# The regional distinctions do not make the relationship between Institutional 
# Performance and Civic Community Index spurious. In Model 1, we observe a strong positive 
# relationship between civic community and institutional performance. When we control for 
# the North-South division in Model 2, the civic community coefficient remains positive 
# and statistically significant, suggesting the relationship is not entirely explained by 
# regional differences. The interactive Model 3 shows that while the strength of the 
# relationship may differ between regions, the fundamental positive association persists 
# in both North and South. Therefore, civic community has an independent effect on 
# institutional performance beyond regional distinctions.
```

# Question 7: Putnam Dataset - Models with Economic Modernization as main predictor

```{r q7}
# Model 1: Simple linear regression (InstPerform ~ EconModern)
econ_model1 <- lm(InstPerform ~ EconModern, data = putnam)

# Model 2: Additive model with EconModern + North
econ_model2 <- lm(InstPerform ~ EconModern + North, data = putnam)

# Model 3: Interactive model with EconModern * North
econ_model3 <- lm(InstPerform ~ EconModern * North, data = putnam)

# Show all three economic modernization models in one table
stargazer(econ_model1, econ_model2, econ_model3, type = "text",
          title = "Regression Models: Institutional Performance and Economic Modernization",
          column.labels = c("Simple", "Additive", "Interactive"),
          dep.var.labels = "Institutional Performance")
```

# Question 8: Graphs for Economic Modernization Models

```{r q8}
# 8a: Graph for Model 1 (EconModern range)
pred_data_e1 <- data.frame(
  EconModern = seq(min(putnam$EconModern, na.rm = TRUE),
                          max(putnam$EconModern, na.rm = TRUE),
                          length.out = 100)
)
pred_data_e1$predicted <- predict(econ_model1, newdata = pred_data_e1)
#Plot marginal effect of Economic Modernization (Model 1)
ggplot(pred_data_e1, aes(x = EconModern, y = predicted)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Marginal Effect of Economic Modernization (Model 1)",
       x = "Economic Modernization", y = "Predicted Institutional Performance") +
  theme_minimal()

# 8b: Graphs for Model 2 (separate for North and South)
pred_data_e2_north <- data.frame(
  EconModern = seq(min(putnam$EconModern, na.rm = TRUE),
                          max(putnam$EconModern, na.rm = TRUE),
                          length.out = 100),
  North = 1
)
pred_data_e2_north$predicted <- predict(econ_model2, newdata = pred_data_e2_north)
# Prediction data for Model 2, North = 0
pred_data_e2_south <- data.frame(
  EconModern = seq(min(putnam$EconModern, na.rm = TRUE),
                          max(putnam$EconModern, na.rm = TRUE),
                          length.out = 100),
  North = 0
)
pred_data_e2_south$predicted <- predict(econ_model2, newdata = pred_data_e2_south)
# Plot marginal effect in Northern regions (Model 2)
ggplot() +
  geom_line(data = pred_data_e2_north, aes(x = EconModern, y = predicted), 
            color = "blue", size = 1) +
  labs(title = "Marginal Effect of Economic Modernization - Northern Regions (Model 2)",
       x = "Economic Modernization", y = "Predicted Institutional Performance") +
  theme_minimal()
# Plot marginal effect in Southern regions (Model 2)
ggplot() +
  geom_line(data = pred_data_e2_south, aes(x = EconModern, y = predicted), 
            color = "red", size = 1) +
  labs(title = "Marginal Effect of Economic Modernization - Southern Regions (Model 2)",
       x = "Economic Modernization", y = "Predicted Institutional Performance") +
  theme_minimal()

# 8c: Graph for Model 3 (two panels)
pred_data_e3 <- expand.grid(
  EconModern = seq(min(putnam$EconModern, na.rm = TRUE),
                          max(putnam$EconModern, na.rm = TRUE),
                          length.out = 100),
  North = c(0, 1)
)
pred_data_e3$predicted <- predict(econ_model3, newdata = pred_data_e3)
# Label region (North vs South) for each row
pred_data_e3$Region <- ifelse(pred_data_e3$North == 1, "North", "South")
# Plot marginal effect by region with two panels (Model 3)
ggplot(pred_data_e3, aes(x = EconModern, y = predicted, color = Region)) +
  geom_line(size = 1) +
  facet_wrap(~Region) +
  labs(title = "Marginal Effect of Economic Modernization by Region (Model 3)",
       x = "Economic Modernization", y = "Predicted Institutional Performance") +
  theme_minimal()
```

# Question 9: Spuriousness Analysis - Economic Modernization

```{r q9}
# The regional distinctions do make the relationship between Institutional 
# Performance and Economic Modernization at least partially spurious. In Model 1, we 
# observe a positive relationship between economic modernization and institutional 
# performance. However, when we control for the North-South division in Model 2, the 
# coefficient for economic modernization decreases substantially and may lose statistical 
# significance, suggesting that much of the apparent effect was due to regional differences. 
# The interactive Model 3 may show different slopes for North and South, indicating that 
# the relationship varies by region. This suggests that regional distinctions confound 
# the relationship between economic modernization and institutional performance, making 
# the initial relationship at least partially spurious.
```

# Question 10: OLS Derivation

```{r q10}
# Derivation of OLS coefficient estimates for Y_i = α + β X_i + μ_i
# 
# The OLS estimator minimizes the sum of squared residuals:
# min Σ(Y_i - α - β X_i)²
# 
# Taking the first-order conditions:
# ∂/∂α Σ(Y_i - α - β X_i)² = 0
# ∂/∂β Σ(Y_i - α - β X_i)² = 0
# 
# Take the derivative with respect to α and set to zero:
# -2Σ(Y_i - α - β X_i) = 0
# ΣY_i - nα - βΣX_i = 0
# α = Ȳ - β X̄
# 
# Take the derivative with respect to β and set to zero:
# -2Σ X_i(Y_i - α - β X_i) = 0
# Σ X_i Y_i - αΣX_i - βΣX_i² = 0
# 
# Substitute α = Ȳ - β X̄ into the second equation:
# Σ X_i Y_i - (Ȳ - β X̄)ΣX_i - βΣX_i² = 0
# Σ X_i Y_i - ȲΣX_i + β X̄ΣX_i - βΣX_i² = 0
# Σ X_i Y_i - nȲX̄ = β(ΣX_i² - nX̄²)
# 
# Therefore the slope estimate is:
# β = Σ(X_i - X̄)(Y_i - Ȳ) / Σ(X_i - X̄)²
# β = Cov(X,Y) / Var(X)
# 
# And the intercept estimate is:
# α = Ȳ - β X̄
```