---
title: "HW Week 10 Jones"
format: pdf
author: "Jones Graham"
---
```{r}
library(ggplot2)
library(tidyverse)
ID <- c(1:6) 
```
# Create numeric vector 1 through 6

```{r}
Race <- c("black", "white", "black", "hispanic", "white", "white")
```
# Create character vector for race

```{r}
class(Race) # Check object type of Race
is.factor(Race) # Check if Race is a factor
is.character(Race) # Check if Race is a character vector
is.logical(Race) # Check if Race is logical
```
## It is a character vector.
```{r}
Voted_For_Obama <- c(TRUE, TRUE, TRUE, FALSE, FALSE, TRUE) 
```
# Create logical vector for Obama vote

```{r}
class(Voted_For_Obama) 
```
# Check type of Voted_For_Obama
```{r}
is.logical(Voted_For_Obama) 
```
# Confirm it is logical
# # It is a logical vector.
```
```{r}
Party_ID <- c("Dem", "Dem", "Dem", "Rep", "Ind", "Rep") 
```
# Create party ID vector

```{r}
Income_Level <- c("High", "Low", "Low", "High", "High", "Low") 
```
# Create income level vector

```{r}
Approval <- c(70, 80, 68, 20, 10, 60) 
```
# Create numeric approval ratings  

```{r}
vote.data <- data.frame(ID, Race, Voted_For_Obama, Party_ID, Income_Level, Approval) 
```
# Combine vectors into data frame

```{r}
is.factor(vote.data$Race)
```
# Check if Race column became a factor in data frame

```{r}
length(vote.data$Voted_For_Obama[vote.data$Voted_For_Obama == TRUE])
```
# Count TRUE (Obama voters)

```{r}
vote.data$Voted_For_Obama[vote.data$Party_ID == "Ind"] 
```
# Check if Independent voted for Obama

```{r}
vote.white <- vote.data[vote.data$Race == "white", ] 
``` 
# Subset data to include only white respondents

```{r}
vote.white[, 3] 
```
# Show 3rd column (Voted_For_Obama) of subset

```{r}
length(vote.white$Voted_For_Obama[vote.white$Voted_For_Obama == TRUE])
```
# Count TRUE in white subset

```{r}
Country_ID <- seq(from = 1, to = 7)
Country_Name <- c("United States", "United Kingdom", "Japan", "China", "Brazil", "Germany", "Egypt")

Regime_Type <- c("Democracy", "Democracy", "Democracy", "Dictatorship",
                 "Democracy", "Democracy", "Dictatorship")

GDP_PC <- c(51163, 39367, 46838, 6070, 11347, 41376, 3115)

EU_Member <- c(FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE)

my.data <- data.frame(Country_ID, Country_Name, Regime_Type, GDP_PC, EU_Member)
```

```{r}
my.data$Population <- c(318946000, 64105654, 127090000, 1367420000, 
                        203322000, 80781000, 87354300)

my.data$GDP <- my.data$GDP_PC * my.data$Population 
# Add population column
``` 

```{r}
my.data$Is_Democracy <- my.data$Regime_Type == "Democracy"  
# Create logical column for democracies
```

```{r}
my.data # Display full data frame
```